Synthesizing Symmetric Lenses

### 2018-07-10 papers/SymmetricLenses/main.tex
add high level alg and main tex

Initially, we hypothesized that extending the synthesis algorithms
of Miltner \ETAL~\cite{optician} to operate over symmetric rather than
bijective lenses
would be relatively straightforward:  Simply replace the bijective
combinators with symmetric ones 
and search using similar heuristics.  However, 
this naive approach runs into two difficulties.

The first difficulty is pragmatic. Symmetric lenses usually operate
over three structures: the source $S$, the target $T$ and a complement
$C$ that contains the information not
present in either $S$ or $T$~\cite{symmetric-lenses}. Logistically, the complement must be stored
somewhere and managed. More significantly, complements complicate synthesis
specifications.
Users can no longer just give examples of source and target
pairings; instead they must also indicate how to manage the 
``hidden state.'' To avoid such complexities, we define a new variant
of symmetric lenses: \emph{simple symmetric lenses}. At the price
of some expressiveness, these new lenses do not require complements.
%only the usual source and target.
%
To characterize their expressive power,
we prove that simple symmetric lenses include all (and only)
those symmetric lenses that satisfy an intuitive property called
\emph{forgetfulness}, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are the complete set of symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $S$ or $T$ when making a round trip. They only need the most
recent instance. 
We justify this loss of expressiveness pragmatically. 
We extend Boomerang~\cite{Boomerang} with simple symmetric lenses and
successfully use them on a range of real-world applications, showing
in the process that simple symmetric lenses integrate smoothly with 
other advanced lenses in Boomerang, including quotient
lenses and matching lenses.

The second difficulty with synthesizing symmetric lenses is more
fundamental: While the number of bijective lenses between two
related formats is often tiny, the number of symmetric lenses is often large. In
practice, the difference in cardinality means that when a search algorithm
selects the first bijective lens it finds, that lens has a good chance of being
the one a user wants. In contrast, and to our surprise, our experiments show
that if our search algorithm selects the first symmetric lens it finds, it
typically fails to find the lens the user wants. The flexibity these lenses
admit when it comes to eliding information generates an explosion of possible
transformations. To disambiguate between the many possible lenses, we need new
principles for identifying the ``likely'' ones, and we need new algorithms to
search the space of program transformations using these principles.

>>>>

Initially, we hypothesized that extending the synthesis algorithms
of Miltner \ETAL~\cite{optician} to operate over symmetric rather than
bijective lenses
would be relatively straightforward:  Simply replace the bijective
combinators with symmetric ones 
and search using similar heuristics.  However, 
this naive approach runs into two difficulties. \dpw{I got into the
weeds a little bit there. 
The goal was to set up the idea that there are ``two parts'' to a symmetric
lens transformation: the information from $s$ and the information from $t$.  In
a future paragraph in the intro we are going to explain that the central heuristic
involves minimizing the information from $t$.}

The first difficulty was pragmatic: symmetric lenses actually usually operate
over three structures, the source $S$, the target $T$ and a complement
$C$~\cite{symmetric-lenses}. The latter object contains all the information not
present in one of $S$ or $T$. Logistically, the complement object must be stored
away somewhere and infrastructure must be developed to manage it. Perhaps more
relevant though, is that complements complicate specifications, even when those
specifications are just examples. The issues is that one cannot just give
examples of source and target; one must also give guidance on how to manage the
``hidden state.'' In order to avoid such complexities, we defined a new variant
of symmetric lenses, which we call \emph{simple symmetric lenses}. At the price
of some expressiveness, our new lenses do not require complements; they only
need source and target data as inputs.

In order to characterize the expressive power of simple symmetric lenses
theoretically, we prove that simple symmetric lenses include all (and only)
those symmetric lenses that satisfy an intuitive property called
\emph{forgetfulness}. In other words, they are the complete set of symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $S$ or $T$ when making a round trip -- they only need the most
recent instance. From a pragmatic perspective, we justify our design choices by
extending the Boomerang implementation~\cite{Boomerang} with simple symmetric
lenses applying them successfully to a range of real-world applications, and
demonstrating that they can enrich, not just with basic bijective combinators,
but with a rich set of advanced features including quotient lenses and matching
lenses.

The second difficulty with synthesizing symmetric lenses is more
fundamental\dpw{better word?}: While the number of bijective lenses between two
related formats is often tiny, the number of symmetric lenses is often large. In
practice, the difference in cardinality means that when a search algorithm
selects the first bijective lens it finds, that lens has a good chance of being
the one a user wants. In contrast, and to our surprise, our experiments show
that if our search algorithm selects the first symmetric lens it finds, it
typically fails to find the lens the user wants. The flexibity these lenses
admit when it comes to eliding information generates an explosion of possible
transformations. To disambiguate between the many possible lenses, we need new
principles for identifying the ``likely'' ones, and we need new algorithms to
search the space of program transformations using these principles.
